### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished single set algorithm description

In this section Brunner and Brantley's fully asynchronous algorithm is
presented along with its application to the Neumann-Ulam method. For
considerably more detail, the algorithm presented can be found in
\citep{brunner_efficient_2009} and was effectively implemented
verbatim for this work. In their work they identify two data sets that
are required to be communicated: the sharing of particles that are
transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Both of these
processes will be addressed at a high level along with how the Monte
Carlo data structures they require are constructed in parallel.

\subsection{Parallel Transport Domain and Source }
\label{subsec:domain_generation}
To utilize a parallel transport algorithm, we must generate the
required data with the correct parallel decomposition. For the
Neumann-Ulam method, the transport domain consists of all states in
the system that are local, and the probabilities and weights for all
state transitions possible in the local domain. At the core of this
representation is the Neumann-Ulam decomposition of the linear
operator. Given an input decomposition for the linear operator, by
definition the Neumann-Ulam decomposition will have the same parallel
decomposition. In addition, any modification made to the linear
operator through preconditioning or relaxation parameters will also
modify the Neumann-Ulam decomposition. All possible transitions are
described by the local graph of the sparse input matrix. In addition,
any left preconditioning or relaxation parameters will modify the
right hand side of the linear system and therefore the fixed source in
the Monte Carlo calculation. This source vector will have the same
parallel decomposition as the input matrix and therefore all of the
birth states for the histories will exist for at least a single
transition event within the local domain.

@@

In this example, we currently reside in state $m$ of the system,
directly correlating to physical location $(i,j)$ in the mesh (the
center node). The discretization stencil of the diffusion problem
dictates the structure of this graph and the states to which a history
may transition. If we play the Monte Carlo game to move to a new state
$n$, then we may move to any of the nodes in this graph, including the
node at which we started. If grid point $(i,j)$ in this example is on
the right of the local domain boundary and we transition to the right
to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
state that is owned by the adjacent domain. We must then gather the
data in the Neumann-Ulam decomposition from the neighboring process
that owns grid point $(i+1,j)$. Doing this provides us with the data
required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
computed locally for this particular boundary transition. For the
parallel source, these adjacent states are not required as histories
may only be born in states in the local domain. Once the local
Neumann-Ulam decomposition has been generated with the proper data
collected from adjacent domains, Monte Carlo transport may proceed in
parallel.

@@

Here we present in detail Brunner and Brantley's 2009 algorithm in
detail and discuss how it is adapted to parallelize the Neumann Ulam
method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
top level sequence performs the task of managing history transport
through the local domain, communication of histories to adjacent
domains, and the completion of transport. For each of these tasks,
additional algorithms, shown in bold in
Algorithm~\ref{alg:parallel_mc_algorithm}
(e.g. \textbf{LocalHistoryTransport}), are presented for additional
detail.

@@

\subsubsection{Domain-to-Domain Communication Kernels}
\label{subsubsec:domain_to_domain_kernel}

@@

@@

\subsubsection{Ending the Transport Sequence}
\label{subsubsec:ending_transport}

@@

>>>>

In this section Brunner and Brantley's fully asynchronous algorithm is
presented along with its application to the Neumann-Ulam method. For
considerably more detail, the algorithm presented can be found in
\citep{brunner_efficient_2009} and was effectively implemented
verbatim for this work. In their work they identify two data sets that
are required to be communicated: the sharing of particles that are
transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Both of these
communication sequences will be addressed at a high level along with
how the Monte Carlo data structures they require are constructed in
parallel.

\subsection{Parallel Transport Domain and Source }
\label{subsec:domain_generation}
To utilize a parallel transport algorithm, we must generate the
required data with the correct parallel decomposition. For the
Neumann-Ulam method, the transport domain consists of all states in
the system that are local, and the probabilities and weights for all
state transitions possible in the local domain. At the core of this
representation is the Neumann-Ulam decomposition of the linear
operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
input decomposition for the linear operator, by definition the
Neumann-Ulam decomposition will have the same parallel
decomposition. In addition, any modification made to the linear
operator through preconditioning or relaxation parameters will also
modify the Neumann-Ulam decomposition. All possible transitions are
described by the local graph of the sparse input matrix. In addition,
any left preconditioning or relaxation parameters will modify the
right hand side of the linear system and therefore the fixed source in
the Monte Carlo calculation. This source vector will have the same
parallel decomposition as the input matrix and therefore all of the
birth states for the histories will exist for at least a single
transition event within the local domain.

@@

In this example, we currently reside in state $m$ of the system,
directly correlating to physical location $(i,j)$ in the mesh (the
center node). The discretization stencil of the diffusion problem
dictates the structure of this graph and the states to which a history
may transition. If we play the Monte Carlo game to move to a new state
$n$, then we may move to any of the nodes in this graph, including the
node at which we started. If grid point $(i,j)$ in this example is on
the right of the local domain boundary and we transition to the right
to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
state that is owned by the adjacent domain. We must then gather the
data in the Neumann-Ulam decomposition from the neighboring process
that owns grid point $(i+1,j)$. Doing this provides us with the data
required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
computed locally for this particular boundary transition. In addition,
we also collect the identification number for the process that owns
this state such that we have an address to send all histories that
leave the local domain boundary for this state. For the parallel
source, these adjacent states are not required as histories may only
be born in states in the local domain. Once the local Neumann-Ulam
decomposition has been generated with the proper data collected from
adjacent domains, Monte Carlo transport may proceed in parallel.

@@

Here we present in detail Brunner and Brantley's 2009 algorithm in
detail and discuss how it is adapted to parallelize the Neumann Ulam
method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
top level sequence performs the task of managing history transport
through the local domain, communication of histories to adjacent
domains, and the completion of transport. For each of these specific
tasks, additional algorithms, shown in bold in
Algorithm~\ref{alg:parallel_mc_algorithm}
(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
detail in the same manner as Brunner and Brantley.

@@

Successful execution of this algorithm requires construction of the
parallel Neumann-Ulam decomposition as described in the preceeding
section. This data is used to begin
Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
collected list of neighboring processors generated while building the
Neumann-Ulam decompostion to setup the set of asynchronous messages
required for domain-to-domain communication. Again, consider this
communication patter for the 9 subdomain example given by
Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each domain
has two neighbors with which they will communicate parallel histories
and this communication goes both ways as represented by the adjacent
arrows in the figure. For each set of neighbors, a non-blocking send
and receive is required with data buffers allocated with a
user-defined size prepared for incoming and outgoing histories. This
nonblocking structure is critical to the performance of the algorithm
in that it permits local history transport to receive while new work
to do is being collected. When a given process is ready to do more
work, it can check these data buffers for incoming histories requiring
further transport. In this way there is a maximum amount of overlap
between communication and transport of histories.

Once the nearest-neighbor communication sequence has been prepared,
the completion of transport sequence is readied in lines 6 and 8-12 in
Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
setting up an asynchronous binary communication tree as presented for
the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
this communication pattern, each process has one parent process
(except for the MASTER process 0) to which it will nonblocking send
the number of histories that terminated their transport procedure by
weight cutoff in the local domain. Equivalently, each process has up
to two child processes from which it will receive their completed
number of histories with a nonblocking receive operation. Setting up a
tree in this manner lets the completed history tally
(\textit{historiesCompleted} in the algorithm) be updated
incrementally and funnelled to the root process. Because we are
solving fixed source problems with the Neumann-Ulam algorithm without
any type of variance reduction that may generate more histories that
we started with, once the root process completion sum tallies to the
number of histories in the source, transport is complete. Once this
occurs, the root process nonblocking sends a process to its children
and each process nonblocking receives a stop message from its parent
as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
then propagated up the tree in the same manner.

With these communication structures prepared, we can now enter the
main transport loop at line 13 of
Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
mutually exclusive tasks enabled by the fully asynchronous
communication patterns are executed until the stop signal is received
from the parent process in the binary tree. For this algorithm, being
mutually exclusive means that all local operations occur independently
of other processes in the system and their current state. For each
process, there are two data structures from which histories may be
obtained for the transport procedure. The first is the local source
and the second is a LIFO (last in first out) stack of histories
transported to the local domain from the adjacent domain. In lines
14-17, if histories exist in either of these data structures, then
they are transported through the local domain using
Algorithm~\ref{alg:local_history_transport}. If the history is
terminated by the weight cutoff the \textit{historiesCompleted} tally
is updated and if it hits the boundary of the domain it is added to
the outgoing history buffer for the neighboring dokmain. If that
buffer is full, it is sent to the neighboring domain with a
nonblocking operation and the memory associated with that buffer
reallocated. In all instances that
Algorithm~\ref{alg:local_history_transport} is executed,
\textit{localProcessed} is incremented as a measure of the number of
histories that have been processed locally.

@@

Continuing in the transport loop, if there are no local histories to
transport or the \textit{localProcessed} count has reached some
user-define check frequency, lines 18-21 in
Algorithm~\ref{alg:local_history_transport} check for more histories
to transport in the incoming data buffers by calling
Algorithm~\ref{alg:process_messages}. For each neighboring domain in
the problem, if there are histories in those data buffers then they
are added to the stack from processing and the nonblocking receive
operation reinstantiated. In addition, the terminated histories count
is updated with those values received from the child processes.

@@

Finally, the transport loop finishes in lines 22-24 of
Algorithm~\ref{alg:local_history_transport} by calling
Algorithm~\ref{alg:control_termination} if there are no local
histories to transport in the stack or the source. In
Algorithm~\ref{alg:control_termination}, we continue to forward the
terminated history tally down to the parent in the binary tree and
send all history buffers to our neighbors, even if they are not
full. If Algorithm~\ref{alg:control_termination} is called by the
MASTER process, it checks for completion of the problem and if it is
complete, forwards to stop flag onto its children as in
Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
then we check for a stop signal from the parent and forward to the
child processes if transport is complete.

@@

In this manner the transport loop continues until all process have
received the stop signal from their parents. Older versions of this
algorithm, particularly some of those presented in
\citep{brunner_comparison_2006}, use a master/slave approach as given
by Figure~\ref{fig:master_comm_tree} for the completion of a transport
stage instead of the binary tree scheme. In this approach, the root
process still manages the final summation of the completion tally, it
directly receives completed tally results from all other processes in
the problem instead of from just its children. Although this
implementation is potentially simpler to understand than the binary
tree approach, the authors of \citep{brunner_comparison_2006} observed
that this type of termination pattern, even when implemented in a
fully asynchronous manner, was a major scalability bottleneck. This
bottleneck is due to the fact that the master process falls behind the
others, creating a load imbalance even with the addition of a few
integer addition operations. We will explicitly demonstrate in scaling
studies how this bottleneck also occurs within a Neumann-Ulam
implementation of this algorithm thus requiring the implementation of
a binary communication tree.
